Avoiding the Global Sort: \break A Faster Contour Tree Algorithm \footnote{The full updated version of this paper is available on the arXiv \cite{rs-mvrms-14}}

### 2016-03-17 paint-contour-tree.tex
changes for camera ready

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%
%%%%%%%%%%Comment out the switch below for full version 
%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%
%\def\SoCGVer{TRUE}%
%%%
\def\HideComments{TRUE}%
%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\newcommand{\CiteFullVer}{\cite{rs-mvrms-14}}%

@@

\InSoCGVer{%
   \documentclass[a4paper,UKenglish]{lipics}
%   \documentclass[11pt]{article}%

@@

@@

%\usepackage{thmtools}
%\usepackage{thm-restate}
%\declaretheorem[name=Theorem,numberwithin=section]{thm}

\newcommand{\atgen}{\symbol{'100}}
\newcommand{\si}[1]{#1}
\newcommand{\BenThanks}[1]{\thanks{%
      Department of Computer Science; %
      University of Illinois; %
%      201 N. Goodwin Avenue; %
      Urbana, IL, USA; %
      {\tt \si{raichel}2\atgen{}\si{uiuc}.\si{edu}}; %
      {\tt \url{\si{http://illinois.edu/\string~\si{raichel2}}}}%
      . %
      #1}}

@@

\Copyright{Benjamin Raichel and Seshadhri Comandur}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

% %Editor-only macros:: begin (do not touch as author)%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
% \serieslogo{}%please provide filename (without suffix)
% \volumeinfo%(easychair interface)
%   {Billy Editor and Bill Editors}% editors
%   {2}% number of editors: 1, 2, ....
%   {Conference title on which this volume is based on}% event
%   {1}% volume
%   {1}% issue
%   {1}% starting page number
% \EventShortName{}
% \DOI{10.4230/LIPIcs.xxx.yyy.p}% to be completed by the volume editor
% % Editor-only macros::end %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
}

@@

All existing algorithms begin with a global sort of at least all critical values of $f$,
which can require (roughly) $\Omega(n\log n)$ time.
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first algorithm whose time complexity depends
on the contour tree structure, and avoids the global sort for non-pathological inputs.
If $C$ denotes the set of critical points in $\MM$, the running time is
roughly $O(\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree.
This matches all existing upper bounds, but is a significant improvement when the contour tree is short and fat.
Specifically, our approach ensures that any comparison made is between nodes in the same descending path in the contour tree,
allowing us to argue strong optimality properties of our algorithm.

@@

Geometric data is often represented as a function $f: \RR^d \to \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some triangulated mesh (i.e.\ simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a topological structure used to represent and visualize
the function $f$. 
% For example, consider a rocket engine during a burn.  The 3D mesh represents the locations of points in space around the engine, 
% and $f$ represents the temperature at each point.  The boundary of the flame is then represented by an isotherm, i.e.\ a contour. 
% 
It is convenient to think of $f$ as a manifold sitting in $\RR^{d+1}$,
%  or alternately consider
% $\MM$ in $\RR^{d+1}$
with the last coordinate (i.e.\ height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components, contours acquire or lose handles. 
The contour tree is a concise representation of all these events.
Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
For $d>2$, some subsequent works, such as \cite{csa-cctad-00}, only include changes in the number of components.

@@

Consider $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
(We assume that $f:\MM \to \RR$ a linear interpolant over distinct valued vertices, where the contour tree
$T$ has maximum degree $3$. The degree assumption simplifies the presentation, and is commonly made~\cite{kobps-ctsssit-97}.)
% \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
%It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-00} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
There is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

@@

\begin{theorem} \label{thm:main-corr} Consider a simplicial complex $f:\MM \to \RR$, described as above, and denote
the contour tree by $T$ with vertex set (the critical points) $C(T)$. There exists an algorithm to compute the contour tree $T$
in $O(\sum_{v \in C(T)} \log \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
compares function values at pairs of points that are ancestor-descendant in $T$.
\end{theorem}

Essentially, the ``run time per critical point" is the height/depth of the point in the contour tree.
This bound immediately yields a run time of $O(t\log D + t\alpha(t) + N)$,
where $D$ is the diameter of the contour tree. 
This is a significant improvement for short and fat contour trees. For example, if the tree is balanced,
then we get a bound of $O(t\log\log t)$.
Even if $T$ contains a long path of length $O(t/\log t)$, but is otherwise short, we get the improved bound of $O(t\log\log t)$.

@@

\begin{theorem} 
\label{thm:main-alg} 
%Consider a PL-Morse\footnote{This is a standard ``well-behaved" input assumption that implies that all critical points have distinct values. 
%Our algorithm does not need this, but it simplifies the proof and presentation.} 
There is a deterministic algorithm to compute the contour tree, $T$, whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
The number of comparisons made is $O(\sum_{p \in P(T)} |p|\log |p| + N)$.
In particular, any comparisons made are only between ancestors and descendants in the contour tree.
\end{theorem}

Note that \Thm{main-corr} is a direct corollary of this statement. For any $v$, $\ell_v$ is at most
the length of the path in $P(T)$ that contains $v$.
This bound is strictly stronger,
since for any balanced contour tree, the run time bound of \Thm{main-alg} is $O(t\alpha(t) + N)$, and $O(t)$
comparisons are made.

The bound of \Thm{main-alg} may seem artificial, since it actually depends on the $P(T)$ that
is implicitly constructed by the algorithm. Nonetheless, we prove that the algorithm of \Thm{main-alg}
has strong optimality properties. For convenience, fix some value of $t$, and consider the set
of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
from $t$ to $t\log t$. Consider some $C \in [t,t\log t]$, and consider the set of terrains
where the algorithm makes $C$ comparisons. Then \emph{any algorithm} must make roughly $C$
comparisons in the worst-case over this set. (All further details are in \InSoCGVer{Appendix~}\Sec{lb}.)

\begin{theorem} 
\label{thm:main-lb}
There exists some absolute constant $\alpha$ such that the following holds.
For sufficiently large $t$ and any $C\in [t, t\log t]$, consider the set $\bF_C$ of terrains with $t$ critical points such that
the number of comparisons made by the algorithm of \Thm{main-alg} on these terrains is in $[C,\alpha C]$.
Any algebraic decision tree that correctly computes
the contour tree on all of $\bF_C$ has a worst case running time of $\Omega(C)$.
\end{theorem}
% 
% \myparagraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
% of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
% characterize hard instances.
% % explain \emph{when}
% % computing contour trees is expensive, and provide insight into the hardness of particular instances. 
% %It is intriguing
% %that the cost (as defined earlier) roughly captures the hardness of an instance.
% Trees that are short and fat are easy to compute, whereas trees with long paths are hard.
% This is tangentially related to the concept of instance-optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
% Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
% of easy to hard instances. In general, our analysis forces one to look closely
% at the relationship between sorting and the contour tree problem. 
%

@@

The first formal result was an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions, by van Kreveld \etal \cite{kobps-ctsssit-97}. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-00} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
% Carr's thesis shows relationships between monotone paths
% in $\MM$ and its contour tree $\cC(\MM)$.

@@

The critical points are exactly where the topology of level sets change.
By assuming that our manifold is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
We abuse notation and refer to this entire set of vertices as a maximum or minimum.

@@

We denote the contour tree of $\MM$
by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
It is not immediately obvious that this graph is a tree, but alternate definitions of the contour tree
in~\cite{csa-cctad-00} imply this is a tree. 
Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.
%
Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
via vertex unfolding (which can increase surface complexity if multi-saddle degrees are allowed to be super-constant).
Therefore, to simplify the presentation, for the remainder of the paper up and down-degrees are at most $2$, and total degree is at most $3$.
%
%We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
%This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).

@@

\InSoCGVer{See \Sec{techmarks} for further technical remarks on the above definitions.}
\newcommand{\technicalRemarks}{
%\subsection{Some technical remarks}
Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
In particular, two $(f(x)+\delta)$-contours of $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
For example, when $d=3$ this is equivalent to the contour gaining a handle.
When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.

@@

Using ideas from~\cite{csa-cctad-00}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}, 
which unlike the contour tree tracks superlevel sets rather than level sets.
% For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
% except at a unique boundary face (think of a large boundary triangle that is the boundary).
Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the connected components of the portion of $\MM$ above height $h$. For a terrain, these superlevel
sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
until finally, we end up with all of $\MM$. The join tree tracks exactly these events.

@@

Of course, things are more complicated when there are other mounds. There may be a yellow mound,
corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In the join tree, $\ell_1$ and $\ell_3$
are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
merge red and blue, until yellow and blue merge. 
A critical insight is the use of \emph{binomial heaps}~\cite{Vu78} to handle all the priorities and the merging.
This ensures that all comparisons are only made between points that are ancestor-descendant in the join tree.
There are numerous details to handle to prove the final run time bound, but this should provide the reader
some intuition behind our algorithm.

@@

\subsection{Breaking $\MM$ into simpler pieces} \label{sec:break}
Let us begin with the algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-00}. The key insight is to build
two different trees, called the join and split trees, and then merge them together into the contour tree.
Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
are the connected components of the portion of $\MM$ above height $h$. For a terrain, the superlevel
sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
until finally, we end up with all of $\MM$. The join tree tracks exactly these events. 
Formally, let $\MM^+_v$ denote the simplicial complex induced on the subset of 
vertices which are higher than $v$.

@@

A major insight of~\cite{csa-cctad-00} is an ingeniously simple linear time procedure to construct
the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e.\ unlike \Fig{sorting}) 
the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-00}
to get the contour tree requires having the split vertices in the join tree. Without
this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.

@@

We can prove that the split tree is just a path, and the contour tree is exactly
the join tree. The formal argument is a little involved, and we employ the merging procedure
of~\cite{csa-cctad-00} to get a proof. We demonstrate that the merging procedure will
actually just output the join tree, so we do not need to actually compute the split tree.
(The real definition of minimum dominant is a little more complicated,
so the contour tree is more than just the join tree. But computationally, it suffices
to construct the join tree.)

We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-00},
one may think that it suffices to design faster algorithms for join trees. But this
cannot give the sort of optimality we hope for. Again, consider \Fig{sorting}. Any algorithm
to construct the true join tree must construct the path of splits, which implies
sorting all of them. It is absolutely necessary to cut $\MM$ into pieces where the cost
of building the join tree can be related to that of building $\cC(\MM)$.

@@

The join tree $\cJ(\NN)$ is a rooted tree with the dominant minimum at the root, and we direct edges downwards (towards the root). So it makes sense to talk
of comparable vs incomparable vertices.
We arrive at the main challenge: how to sort only the comparable critical points, without
constructing the join tree? The join tree algorithm of~\cite{csa-cctad-00} is a typical event-based
computational geometry algorithm. We have to step away from this viewpoint to avoid the global sort.

@@

\InSoCGVer{The following theorems are intuitively obvious, and are proven in \Sec{proofOfSurgery}.}

@@

\begin{definition} \label{def:dom} A simplicial complex is \emph{minimum dominant} if there exists
a minimum $x$ such that every non-minimal \emph{vertex} in the manifold has a non-ascending path to $x$.
Analogously define \emph{maximum dominant}. 
\end{definition}

The first aspect of the partitioning is ``raining''. Start at some point $x \in \MM$ and imagine rain at $x$.
The water will flow downwards along non-ascending paths and ``wet'' all the points encountered. Note that this procedure considers all points
of the manifold, not just vertices.

@@

\section{Contour trees of extremum dominant manifolds} \label{sec:extreme}

>>>>

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%
%%%%%%%%%%Comment out the switch below for full version 
%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%
\def\SoCGVer{TRUE}%
%%%
\def\HideComments{TRUE}%
%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

@@

\newcommand{\CiteFullVer}{\cite{rs-mvrms-14}\xspace}%

@@

\InSoCGVer{%
   \documentclass[a4paper,UKenglish]{lipics-v2016}
%   \documentclass[11pt]{article}%

@@

\newcommand{\thmConst}{\gamma}

@@

\newcommand{\atgen}{\symbol{'100}}
\newcommand{\si}[1]{#1}
\newcommand{\BenThanks}[1]{\thanks{%
      Department of Computer Science; %
      University of Texas at Dallas; %
      Richardson, TX, USA; %
      {\tt \si{benjamin.raichel}@\si{utdallas}.\si{edu}}; %
      {\tt \url{\si{http://utdallas.edu/\string~\si{benjamin.raichel}}}}%
      . %
      #1}}

@@

\Copyright{Benjamin Raichel and C. Seshadhri}%mandatory, please use full first names. LIPIcs license is "CC-BY";  http://creativecommons.org/licenses/by/3.0/

@@

\EventEditors{S\'andor Fekete and Anna Lubiw}
\EventNoEds{2}
\EventLongTitle{32nd International Symposium on Computational Geometry
(SoCG 2016)}
\EventShortTitle{SoCG 2016}
\EventAcronym{SoCG}
\EventYear{2016}
\EventDate{June 14-18, 2016}
\EventLocation{Boston, USA}
\EventLogo{}
\SeriesVolume{51}
\ArticleNo{57}

}

@@

All existing algorithms begin with a global sort of at least all critical values of $f$,
which can require (roughly) $\Omega(n\log n)$ time.
Existing lower bounds show that there are pathological instances where this sort is required.
We present the first algorithm whose time complexity depends
on the contour tree structure, and avoids the global sort for non-pathological inputs.
If $C$ denotes the set of critical points in $\MM$, the running time is
roughly $O(\sum_{v \in C} \log \ell_v)$, where $\ell_v$ is the depth of $v$ in the contour tree.
This matches all existing upper bounds, but is a significant asymptotic improvement when the contour tree is short and fat.
Specifically, our approach ensures that any comparison made is between nodes that are either adjacent in $\MM$ 
or in the same descending path in the contour tree,
allowing us to argue strong optimality properties of our algorithm.

@@

Geometric data is often represented as a function $f: \RR^d \to \RR$. Typically, a finite representation is given
by considering $f$ to be piecewise linear over some triangulated mesh (i.e.\ simplicial complex) $\MM$ in $\RR^d$.
\emph{Contour trees} are a topological structure used to represent and visualize
the function $f$. 
It is convenient to think of $f$ as a simplicial complex sitting in $\RR^{d+1}$,
with the last coordinate (i.e.\ height) given by $f$. Imagine sweeping the hyperplane $x_{d+1} = h$
with $h$ going from $+\infty$ to $-\infty$. At every instance, the intersection of this plane 
with $f$ gives a set of connected components, the \emph{contours} at height $h$. As the sweeping 
proceeds various events occur: new contours are created or destroyed, contours merge into each other or
split into new components, contours acquire or lose handles. 
The contour tree is a concise representation of all these events.
Throughout we follow the definition of contour trees from \cite{kobps-ctsssit-97} which includes all changes in topology.  
For $d>2$, some subsequent works, such as \cite{csa-cctad-03}, only include changes in the number of components.

@@

Consider $f: \MM \to \RR$, where $\MM$ is a triangulated mesh with $n$ vertices, $N$ faces in total, and $t \leq n$ critical points. 
(We assume that $f:\MM \to \RR$ a linear interpolant over distinct valued vertices, where the contour tree
$T$ has maximum degree $3$. The degree assumption simplifies the presentation, and is commonly made~\cite{kobps-ctsssit-97}.)
% \Ben{It is standard to assume that $f$ has distinct values on the vertices of $\MM$ and the value at interior points is linearly interpolated.}
%It is standard to assume that $f$ is PL-Morse (a well-behaved input assumption) and that the maximum degree in the contour tree is $3$. 
A fundamental result in this area is the algorithm of Carr, Snoeyink, and Axen to compute contour trees,
which runs in $O(n\log n + N\alpha(N))$ time~\cite{csa-cctad-03} (where $\alpha(\cdot)$ denotes the inverse
Ackermann function). In practical applications, $N$ is typically $\Theta(n)$
(certainly true for $d=2$). The most expensive operation is an initial sort of all the vertex
heights. Chiang \etal build on this approach to get a faster algorithm that only sorts the critical vertices,
yielding a running time of $O(t\log t + N)$~\cite{cllr-sooscctmp-05}.
Common applications for contour trees involve turbulent combustion or noisy data, where the number of critical points is likely to be $\Omega(n)$.
There is a worst-case
lower bound of $\Omega(t\log t)$ by Chiang \etal \cite{cllr-sooscctmp-05}, based on a construction of Bajaj \etal \cite{BaKr+98}.

@@

\begin{theorem} \label{thm:main-corr} Consider a simplicial complex $f:\MM \to \RR$, described as above, and denote
the contour tree by $T$ with vertex set (the critical points) $C(T)$. There exists an algorithm to compute $T$
in $O(\sum_{v \in C(T)} \log \ell_v + t\alpha(t) + N)$ time. Moreover, this algorithm only
compares function values at pairs of vertices that are ancestor-descendant in $T$ or adjacent in $\MM$.
\end{theorem}

Essentially, the ``run time per critical point" is the height/depth of the point in the contour tree.
This bound immediately yields a run time of $O(t\log D + t\alpha(t) + N)$,
where $D$ is the diameter of the contour tree. 
This is a significant asymptotic improvement for short and fat contour trees. For example, if the tree is balanced,
then we get a bound of $O(t\log\log t)$.
Even if $T$ contains a long path of length $O(t/\log t)$, but is otherwise short, we get the improved bound of $O(t\log\log t)$.

@@

\begin{theorem} 
\label{thm:main-alg} 
%Consider a PL-Morse\footnote{This is a standard ``well-behaved" input assumption that implies that all critical points have distinct values. 
%Our algorithm does not need this, but it simplifies the proof and presentation.} 
There is a deterministic algorithm to compute the contour tree, $T$, whose running time is $O(\sum_{p \in P(T)} |p|\log |p| + t\alpha(t) + N)$,
where $P(T)$ is a specific path decomposition (constructed implicitly by the algorithm).
The number of comparisons made is $O(\sum_{p \in P(T)} |p|\log |p| + N)$.
In particular, comparisons are only made between pairs of vertices that are ancestor-descendant in $T$ or adjacent in $\MM$.
\end{theorem}

Note that \Thm{main-corr} is a direct corollary of this statement. For any $v$, $\ell_v$ is at least
the length of the path in $P(T)$ that contains $v$.
This bound is strictly stronger,
since for any balanced contour tree, the run time bound of \Thm{main-alg} is $O(t\alpha(t) + N)$, and $O(N)$
comparisons are made. (Since for a balanced tree, one can show $\sum_{p \in P(T)} |p|\log |p|= O(t)$.)

The bound of \Thm{main-alg} may seem artificial, since it actually depends on the $P(T)$ that
is implicitly constructed by the algorithm. Nonetheless, we prove that the algorithm of \Thm{main-alg}
has strong optimality properties. For convenience, fix some value $t=\Omega(N)$, and consider the set
of terrains ($d=2$) with $t$ critical points. The bound of \Thm{main-alg} takes values ranging
from $t$ to $t\log t$. Consider some $\thmConst \in [t,t\log t]$, and consider the set of terrains
where the algorithm makes $\thmConst$ comparisons. Then \emph{any algorithm} must make roughly $\thmConst$
comparisons in the worst-case over this set. (Further details are in \InSoCGVer{the full version \CiteFullVer}\InNotSoCGVer{\Sec{lb}}.)

\begin{theorem} 
\label{thm:main-lb}
There exists some absolute constant $\alpha$ such that the following holds.
For sufficiently large $t$ and any $\thmConst\in [t, t\log t]$, consider the set $\bF_\thmConst$ of terrains with $t$ critical points such that
the number of comparisons made by the algorithm of \Thm{main-alg} on these terrains is in $[\thmConst,\alpha \thmConst]$.
Any algebraic decision tree that correctly computes
the contour tree on all of $\bF_\thmConst$ has a worst case running time of $\Omega(\thmConst)$.
\end{theorem}
% 
% \myparagraph{Beyond worst-case analysis.} Our result provides understanding of the complexity
% of contour tree construction beyond the standard worst-case analysis. Our theorem and lower bound 
% characterize hard instances.
% % explain \emph{when}
% % computing contour trees is expensive, and provide insight into the hardness of particular instances. 
% %It is intriguing
% %that the cost (as defined earlier) roughly captures the hardness of an instance.
% Trees that are short and fat are easy to compute, whereas trees with long paths are hard.
% This is tangentially related to the concept of instance-optimal algorithms, as defined by Afshani \etal \cite{abc-ioga-09}.
% Their notion of optimality is far more refined that what we prove, but it shares the flavor of understanding the spectrum
% of easy to hard instances. In general, our analysis forces one to look closely
% at the relationship between sorting and the contour tree problem. 
%

@@

The first formal result was an $O(N\log N)$ time algorithm for functions over 2D 
meshes and an $O(N^2)$ algorithm for higher dimensions, by van Kreveld \etal \cite{kobps-ctsssit-97}. 
Tarasov and Vyalya \cite{tv-cct-98} improved the running time to $O(N\log N)$ for the 3D case.
The influential paper of Carr \etal \cite{csa-cctad-03} improved the running time for all dimensions to $O(n\log n + N\alpha(N))$.
Pascucci and Cole-McLaughlin \cite{pc-ectls-02} provided an $O(n+t\log n)$ time algorithm for 
$3$-dimensional structured meshes. Chiang \etal \cite{cllr-sooscctmp-05} provide an unconditional $O(N+t\log t)$ algorithm.
% Carr's thesis shows relationships between monotone paths
% in $\MM$ and its contour tree $\cC(\MM)$.

@@

The critical points are exactly where the topology of level sets change.
By assuming that our simplicial complex is boundary critical, the vertices on a given boundary are either collectively all maxima or all minima.  
We abuse notation and refer to this entire set of vertices as a maximum or minimum.

@@

We denote the contour tree of $\MM$
by $\reeb(\MM)$. The corresponding node and edge sets are denoted as $\cV(\cdot)$ and $\cE(\cdot)$.
It is not immediately obvious that this graph is a tree, but alternate definitions of the contour tree
in~\cite{csa-cctad-03} imply this is a tree. 
Since this tree has height values associated with the vertices, we can talk about up-degrees and down-degrees in $\reeb(\MM)$.
%
Similar to \cite{kobps-ctsssit-97} (among others), multi-saddles are treated as a set of ordinary saddles, which can be realized 
via vertex unfolding (which can increase surface complexity if multi-saddle degrees are allowed to be super-constant).
Therefore, to simplify the presentation, for the remainder of the paper up and down-degrees are at most $2$, and total degree is at most $3$.
%
%We assume there are no ``multi-saddles", so up and down-degrees are at most $2$, and the total degree is at most $3$. 
%This is again a standard assumption in topological algorithms and can be achieved by vertex unfolding (Section 6.3 in~\cite{HaEd10}).

@@

\InSoCGVer{See the full version \CiteFullVer for further technical remarks on the above definitions.}
\newcommand{\technicalRemarks}{
%\subsection{Some technical remarks}
Note that if one intersects $\MM$ with a given ball $B$, then a single contour in $\MM$ might be split into more than one contour in the intersection. 
In particular, two $(f(x)+\delta)$-contours of $f|_{B_\eps(x)}$, given by \Def{deg}, might actually be the same contour in $\MM$. 
Alternatively, one can define the up-degree (as opposed to \emph{Morse} up-degree) as the number of $(f(x)+\delta)$-contours (in the full $\MM$)
that intersect $B_\eps(x)$, a potentially smaller number. This up-degree
is exactly the up-degree of $x$ in $\reeb(\MM)$. (Analogously, for down-degree.)
When the Morse up-degree is $2$ but the up-degree is $1$, the topology of the level set changes but not by the number of connected components changing.
For example, when $d=3$ this is equivalent to the contour gaining a handle.
When $d=2$, this distinction is not necessary, since any point with Morse degree strictly greater than $1$ will have degree strictly greater than $1$ in $\reeb(\MM)$.

@@

Using ideas from~\cite{csa-cctad-03}, we prove that the contour tree of an extremum dominant complex is (basically) just the \emph{join tree}, 
which unlike the contour tree tracks superlevel sets rather than level sets.
% For the $d=2$ case, it helps to visualize such an input as a terrain with no minima,
% except at a unique boundary face (think of a large boundary triangle that is the boundary).
Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the connected components of the portion of $\MM$ above height $h$. For a terrain, these superlevel
sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
until finally, we end up with all of $\MM$. The join tree tracks exactly these events.

@@

Of course, things are more complicated when there are other mounds. There may be a yellow mound,
corresponding to $\ell_3$ that joins with the blue mound higher up at some vertex $u$ (see the right part of \Fig{colors}). In the join tree, $\ell_1$ and $\ell_3$
are sibling leaves, and $\ell_2$ is a sibling of some ancestor of these leaves. So we cannot
merge red and blue, until yellow and blue merge. 
A critical insight is the use of \emph{binomial heaps}~\cite{Vu78}, or any other efficiently mergeable heaps, to handle all the priorities and the merging.
This ensures that all comparisons are only made between points that are ancestor-descendant in the join tree or adjacent in $\MM$.
There are numerous details to handle to prove the final run time bound, but this should provide the reader
some intuition behind our algorithm.

@@

\subsection{Breaking $\MM$ into simpler pieces} \label{sec:break}
Let us begin with the algorithm of Carr, Snoeyink, and Axen~\cite{csa-cctad-03}. The key insight is to build
two different trees, called the join and split trees, and then merge them together into the contour tree.
Consider sweeping down the hyperplane $x_{d+1} = h$ and taking the \emph{superlevel} sets. These 
are the connected components of the portion of $\MM$ above height $h$. For a terrain, the superlevel
sets are a collection of ``mounds". As we sweep downwards, these mounds keep joining each other,
until finally, we end up with all of $\MM$. The join tree tracks exactly these events. 
Formally, let $\MM^+_v$ denote the simplicial complex induced on the subset of 
vertices which are higher than $v$.

@@

A major insight of~\cite{csa-cctad-03} is an ingeniously simple linear time procedure to construct
the contour tree from the join and split trees. So the bottleneck is computing these trees. Observe
in \Fig{sorting} that the split vertices form a long path in the join tree (and vice versa). Therefore, constructing
these trees forces a global sort of the splits, an unnecessary computation for the contour tree. Unfortunately, in general (i.e.\ unlike \Fig{sorting}) 
the heights of joins and splits may be interleaved in a complex manner, and hence the final merging of~\cite{csa-cctad-03}
to get the contour tree requires having the split vertices in the join tree. Without
this, it is not clear how to get a consistent view of both joins and splits, required for the contour tree.

@@

We can prove that the split tree is just a path, and the contour tree is exactly
the join tree. The formal argument is a little involved, and we employ the merging procedure
of~\cite{csa-cctad-03} to get a proof. We demonstrate that the merging procedure will
actually just output the join tree, so we do not need to actually compute the split tree.
(The real definition of minimum dominant is a little more complicated,
so the contour tree is more than just the join tree. But computationally, it suffices
to construct the join tree.)

We stress the importance of this step for our approach. Given the algorithm of~\cite{csa-cctad-03},
one may think that it suffices to design faster algorithms for join trees. But this
cannot give the sort of optimality we hope for. Again, consider \Fig{sorting}. Any algorithm
to construct the true join tree must construct the path of splits, which implies
sorting all of them. It is absolutely necessary to cut $\MM$ into pieces where the cost
of building the join tree can be related to that of building $\cC(\MM)$.

@@

The join tree $\cJ(\NN)$ is a rooted tree with the dominant minimum at the root, and we direct edges downwards (towards the root). So it makes sense to talk
of comparable vs incomparable vertices.
We arrive at the main challenge: how to sort only the comparable critical points, without
constructing the join tree? The join tree algorithm of~\cite{csa-cctad-03} is a typical event-based
computational geometry algorithm. We have to step away from this viewpoint to avoid the global sort.

@@

\InSoCGVer{The following theorems are intuitively obvious, and are proven in the full version \CiteFullVer.}

@@

\begin{definition} \label{def:dom} A simplicial complex is \emph{minimum dominant} if there exists
a minimum $x$ such that every non-minimal \emph{vertex} in the complex has a non-ascending path to $x$.
Analogously define \emph{maximum dominant}. 
\end{definition}

The first aspect of the partitioning is ``raining''. Start at some point $x \in \MM$ and imagine rain at $x$.
The water will flow downwards along non-ascending paths and ``wet'' all the points encountered. Note that this procedure considers all points
of the complex, not just vertices.

@@

\section{Contour trees of extremum dominant complexes} \label{sec:extreme}

The previous section allows us to restrict attention to extremum dominant complexes.
We will orient so that the extremum in question is always a \emph{minimum}.
We will fix such a simplicial complex $\MM$, with the dominant minimum $m^*$. 
%The set of non-dominant minima is denoted by $M$.
For vertex $v$, we use $\MM^+_v$ to denote the simplicial complex obtained by only
keeping vertices $u$ such that $f(u) > f(v)$. Analogously, define $\MM^-_v$. Note that $\MM^+_v$ may contain
numerous connected components.
